---
course: "Bayesian Computational Methods"
assignment: "Tutorial 1"
student-number: "LTHISA001"
format:
  pdf:
    colorlinks: true
    linkcolor: blue
    urlcolor: blue
    documentclass: article
    fontsize: 12pt
    geometry:
      - left=3cm
      - right=3cm
      - top=2.5cm
      - bottom=2.5cm
    toc: false
    toc-depth: 3
    number-sections: false
    listings: true
    include-in-header: 
      - metadata.tex
      - header.tex
    include-before-body: title.tex
execute:
  warning: false
  message: false
editor: visual
---

```{r}
#| echo: false

# Load data: 
require(ElemStatLearn)
data('SAheart')
# View(SAheart)

# Random training data set:
set.seed(1)
training = SAheart[sample(1:462,size=300),]
```

```{r}
#| echo: false

# load required libraries:
library(betareg)
library(ggplot2)
library(dplyr)
library(purrr)
```

# Introduction

We are fitting the following Bayesian logistic regression model to the `SAheart` dataset from the `ElemStatLearn` R package:

$$
log(\frac{p_{i}}{1-p_{i}}) = \beta_0 + \beta_1\text{age} + \beta_2\text{famhist} + \beta_3\text{tobacco},
$$

where

$$
p_i = \text{Pr}(chd_i = 1 \mid \bs{x}_i,\bs{\beta}) \quad \text{and} \quad \bs{x}_i = (1, \text{age}_i, \text{famhist}_i, \text{tobacco}_i)
$$

`famhist` is coded as a binary predictor where $0$ indicates absence and $1$ indicates presence.

# Question a

Expert 1 provided the median probability that a patient would develop coronary heart disease. Based on the methods laid out in Choy et al. (2009), we denote the median probabilities as $\text{Z}_{\text{m}}$ and model them with the following Beta distribution:

$$
\text{Z}_{m} \sim Beta(\mu_m,\gamma_m) \quad \text{for} \quad m = 1, \dots,\text{M} ,
$$

where $\mu_m$ is the mean and $\gamma_m$ is the precision parameter of $\text{Z}_{\text{m}}$. The logit link below relates the expert's elicited median probabilities to the regression coefficients.

$$
logit(\mu_{m}) = \bs{x}_{m}^{T}\beta, \quad \bs{x}^T_m = (1, \text{age}_m, \text{famhist}_m, \text{tobacco}_m).
$$

Under the assumption that the prior distribution of the regression coefficients are multivariate normal random variables:

$$
\bs{\beta} = (\beta_0, \beta_1, \beta_2, \beta_3)^T\sim N_{4}(\bs{\mu}_0,\bs{\Sigma}_0).
$$

$\bs{\mu}_0$ and $\Sigma_0$ can be estimated by fitting a Beta regression model. `betareg`, with a default logit link function was fitted to the 14 elicited median probabilities using covariates `age`, `famhist` and `tobacco`.

Maximum Likelihood Estimators are asymptotically multivariate normal. The joint sampling distribution of the estimated regression coefficients, $\hat\beta$, approaches a Multivariate Normal distribution with mean equal to the true parameter vector. Following the Choy et al. (2009) elicitation procedure, we use the following sample estimates to specify the prior distribution parameters:

$$
\bs{\mu}_0 = \hat{\bs{\beta}} \quad \text{and} \quad \bs{\Sigma}_0 = \widehat{ Var}(\hat{\bs{\beta}}).
$$

Extracting only the first 4 rows and columns from $\widehat{Var}(\hat{\bs{\beta}})$ isolates the regression parameters from the nuisance parameter, $\phi$.

```{r}
#| echo: false

# Table 1: Design points 
age     = c(20,29,32,34,40,41,42,43,46,54,55,61,64,64)
famhist = c(0,0,0,0,0,1,0,0,0,1,1,1,1,1)
tobacco = c(0.00,1.50,0.00,0.40,1.62,0.90,2.92,6.75,
            12.00,7.60,4.80,25.01,8.60,12.18)
E1_med  = c(0.05,0.11,0.1,0.12,0.165,0.33,0.2,0.275,
            0.46,0.68,0.64,0.96,0.8,0.86)

# Store data in a dataframe: 
E1_df = data.frame(
   Z       = E1_med,
   age     = age,
   famhist = famhist,
   tobacco = tobacco
)

# Fit a beta regression model: 
betareg_fit = betareg(Z~age+famhist+tobacco,data=E1_df)


# Obtain mu0:
mu0_a = coef(betareg_fit,model='mean')

# Obtain Sigma0:
S0_a = vcov(betareg_fit)[1:4,1:4]

# Check correlations: 
cor_mat_a = cov2cor(S0_a)
```

# Question b

Expert 2 provided lower and upper quartile values associated with the probability that a patient would develop coronary heart disease, across M=14 design points.

Let $p_m= \text{Pr}(chd_m = 1 \mid \bs{x}_m,\bs{\beta})$, then:

$$
\text{Pr}(\theta_m \le q_{0.25,m}) = 0.25, \quad \text{Pr}(\theta_m \le q_{0.75,m}) = 0.75.
$$

Following Hosack et al. (2017), the 25% and 75% quantiles are transformed to linear predictors via the logit link function:

$$
\eta_{0.25,m} = \text{logit}(q_{0.25,m}) \quad \text{and} \quad \eta_{0.75,m} = \text{logit}(q_{0.75,m}).
$$

Assuming a normal prior for $\bs{\beta}$:

$$
\eta_m = \bs{x}_m^T\beta \sim N(m_m,v_m)
$$

This gives:

$$
\text{Pr}(g(p_m) \le g(q_{0.25,m})) = \Phi(\eta_m;m_m,v_m) = 0.25.
$$

Assuming that $\eta_m = \bs{x}^T_m\bs{\beta} \sim N(m_m,v_m)$, the target interval probabilities are $\bs{l} = (l_1,l_2,l_3)^T = (0.25,0.50,0.25)^T$. For each design point $m$, the parameters $(m_m,v_m)$ are estimated to minimise the Kullback-Leibler (KL) divergence using `nlm()`:

$$
\text{KL}(m_m,v_m) = \sum^{3}_{j=1} \text{log} \Big( \frac{l_j}{Z_j(m_m,v_m)}\Big) l_j.
$$

where

$$
\begin{aligned}
Z_1(m_m, v_m) &= \Phi\left(\frac{\eta_{0.25, m} - m_m}{\sqrt{v_m}}\right), \\[4pt]
Z_2(m_m, v_m) &= \Phi\left(\frac{\eta_{0.75, m} - m_m}{\sqrt{v_m}}\right) - Z_1(m_m, v_m), \\[4pt]
Z_3(m_m, v_m) &= 1 - \Phi\left(\frac{\eta_{0.75, m} - m_m}{\sqrt{v_m}}\right).
\end{aligned}
$$

```{r}
#| echo: false

# Table 1: Design points 
E2_25 = c(0.05,0.08,0.08,0.10,0.14,0.32,0.18,0.25,0.39,0.65,0.59,0.93,0.77,0.82)
E2_75 = c(0.075,0.13,0.125,0.13,0.18,0.39,0.22,0.31,0.5,0.74,0.68,0.98,0.83,0.9)

# No. observations: 
n = length(E2_25)
```

```{r}
#| echo: false

# Define logit function:
logit_fxn = function(x) {
  log(x/(1-x))
}

# Transform fractiles:
p25 = logit_fxn(E2_25)
p75 = logit_fxn(E2_75)

# Kullback-Leibler function:
# This what we wish to optimize 
KL_fxn = function(par,p25,p75) {
  # Mean and standard deviation: 
  m = par[1]
  v = par[2]
  
  # Calculate integrals: 
  Z1 = pnorm(q=p25,mean=m,sd=v)
  Z2 = pnorm(q=p75,mean=m,sd=v)-Z1
  Z3 = 1-pnorm(q=p75,mean=m,sd=v)
  
  # Calculate Kullback-Leibler for each 'integral':
  KL1 = sum(log(0.25/Z1)*0.25)
  KL2 = sum(log(0.50/Z2)*0.50)
  KL3 = sum(log(0.25/Z3)*0.25)
  
  # Return a single Kullback-Leibler value: 
  KL = sum(KL1,KL2,KL3)
  return(KL)
}

m_vec = numeric(n)
v_vec = numeric(n) 

# For loop to optimize m and v for each observation: 
for (i in 1:14) {
  init_m = 0
  init_v = 1
  
  nlm_fit = nlm(f   =KL_fxn,
                p   =c(init_m,init_v),
                p25 = p25[i],
                p75 = p75[i]
                )
  
  m_vec[i] = nlm_fit$estimate[1]
  v_vec[i] = nlm_fit$estimate[2]
}

# m_vec
# v_vec
```

```{r}
#| echo: false

# Design matrix:
X = cbind(1,
          age,
          famhist,
          tobacco)

m = m_vec
V = diag(v_vec)

mu0_b = solve(t(X)%*%solve(V)%*%X)%*%t(X)%*%solve(V)%*%m

S0_b = solve(t(X)%*%solve(V)%*%X)
```

Let $\bs{m} = (m_1, \dots, m_M)^\top$ and $\bs{V} = \text{diag}(v_1, \dots, v_M)$. The prior mean vector $\bs{\mu}_0$ and covariance matrix $\bs{\Sigma}_0$ for Expert 2 are obtained as follows:

$$\bs{\mu}_0 = \left(\bs{X}^\top \bs{V}^{-1} \bs{X}\right)^{-1} \bs{X}^\top \bs{V}^{-1} \bs{m} \quad \text{and} \quad \bs{\Sigma}_0 = \left(\bs{X}^\top \bs{V}^{-1} \bs{X}\right)^{-1}.$$

# Question c

```{r}
#| echo: false 

param_names = c('Intercept','age','famhist','tobacco')

# Prior standard deviations: 
sd_1 = sqrt(diag(S0_a))
sd_2 = sqrt(diag(S0_b))

# Prior comparison:
prior_comparison = data.frame(
  Parameter    = param_names,
  Expert1_Mean = as.numeric(mu0_a),
  Expert2_Mean = as.numeric(mu0_b),
  Expert1_SD   = sd_1,
  Expert2_SD   = sd_2
)
```

```{r}
#| label: tbl-pc
#| tbl-cap: 'Elicited prior mean and standard deviations from expert 1 and expert 2.'
#| echo: false

# Display table: 
prior_comparison = as.data.frame(round(prior_comparison[, -1],4))
colnames(prior_comparison) = c('Expert 1 mean', 
                               'Expert 2 mean', 
                               'Expert 1 sd', 
                               'Expert 2 sd')
rownames(prior_comparison) = param_names

knitr::kable(
  prior_comparison
)

```

The $\beta$ coefficients, in @tbl-pc, corresponding to `age`, `famhist` and `tobacco` for Expert 1 are all positive. A one-unit increase in `age` multiplicatively increases the odds of coronary heart disease by 5.3%. Similarly, a one-unit increase in `tobacco` multiplicatively increases the odds of coronary heart disease by 11.7%. Changing `famhist` from \textit{Absence} to \textit{Presence} multiplies the odds by 2.71. The elicited prior implies a baseline probability of approximately 2%.

\newpage 

Do we need this? 

```{r}
#| label: tbl-S0
#| tbl-cap: 'Elicited prior covariance matrix.'
#| echo: false

# Table of Sigma0 results: 
S0_a_df = as.data.frame(round(S0_a,5))
rownames(S0_a_df) = c('Intercept','age','famhist','tobacco')
colnames(S0_a_df) = c('Intercept','age','famhist','tobacco')
knitr::kable(
  S0_a_df
)
```

The diagonal elements in @tbl-S0 give the marginal variances of the prior regression coefficients. The small values indicate low prior uncertainty.

```{r}
#| label: tbl-cor
#| tbl-cap: 'Prior parameter correlation matrix.'
#| echo: false
# Table of correlation results: 
cor_a_df = as.data.frame(round(cor_mat_a,3))
rownames(cor_a_df) = c('Intercept','age','famhist','tobacco')
colnames(cor_a_df) = c('Intercept','age','famhist','tobacco')
knitr::kable(
  cor_a_df
)
```

The correlation matrix dispalyed in @tbl-cor indicate dependence between the regression coefficients. This elicited prior cannot be assumed to represent independent priors for each regression coefficient but rather a joint multivariate distribution.


discuss whether expert 2 produces a:

-   more concentrated/diffused prior

-   strong/weaker association for each predictor

-   different correlation structure

What does the prior actually believe?

For example:

-   Is the prior strongly concentrated around a particular age effect?

-   Does the expert consider family history important?

-   How uncertain is the tobacco effect?

-   Are the coefficients strongly correlated?

-   Does the prior imply plausible low/high-risk probabilities at the supplied design points?


The two elicitation methods result in different prior means and levels of uncertainty for the regression coefficients. Differences in the prior standard deviations indicate that the experts' probability assessments imply different degrees of certainty, while differences in the prior correlations indicate different beliefs about the dependence between regression effects


```{r}
#| label: fig-prior-comparison
#| fig-cap: "Comparison of marginal prior distributions for Expert 1 and Expert 2."
#| fig-width: 7
#| fig-height: 5
#| echo: false
#| message: false
#| warning: false

mu_E1 = as.vector(mu0_a)
sd_E1 = sqrt(diag(S0_a))
mu_E2 = as.vector(mu0_b)
sd_E2 = sqrt(diag(S0_b))

# Generate density evaluation data for each parameter:
plot_data = map_dfr(1:4, function(i) {
  # Define bounds: 
  x_min  = min(mu_E1[i]-4*sd_E1[i],mu_E2[i]-4*sd_E2[i])
  x_max  = max(mu_E1[i]+4*sd_E1[i],mu_E2[i]+4*sd_E2[i])
  x_grid = seq(x_min,x_max,length.out=300)
  
  tibble(
    Parameter = factor(param_names[i], 
                       levels=param_names),
    x = rep(x_grid, 2),
    Density = c(dnorm(x_grid,mean=mu_E1[i],sd=sd_E1[i]),
                dnorm(x_grid,mean=mu_E2[i],sd=sd_E2[i])),
    Expert = rep(c("Expert 1", "Expert 2"), each = 300)
  )
})

# Plot overlaid density curves: 
ggplot(plot_data, 
       mapping=aes(x = x,
                   y = Density, 
                   color = Expert, 
                   fill = Expert)) +
  geom_line(linewidth = 0.8) +
  geom_area(alpha = 0.15, 
            position = "identity") +
  facet_wrap(~ Parameter, 
             scales = "free", 
             ncol = 2) +
  scale_color_manual(values = c("Expert 1" = "steelblue", 
                                "Expert 2" = "deeppink")) +
  scale_fill_manual(values = c("Expert 1" = "steelblue", 
                               "Expert 2" = "deeppink")) +
  labs(
    x = expression(beta),
    y = "Prior Density",
    color = "Prior Source",
    fill = "Prior Source"
  ) +
  theme_bw(base_size = 10) +
  theme(
    legend.position = "bottom",
    strip.background = element_rect(fill = "gray95"),
    strip.text = element_text(face = "bold")
  )
```

*insert some explanation here* 

\newpage 

# Appendix 

```{r}

```

