---
course: "Time Series"
assignment: "Journal Entries"
student-number: "CMPROX001"
format:
  pdf:
    colorlinks: true
    linkcolor: blue
    urlcolor: blue
    documentclass: article
    fontsize: 12pt
    geometry:
      - left=3cm
      - right=3cm
      - top=2.5cm
      - bottom=2.5cm
    toc: false
    toc-depth: 3
    number-sections: false
    listings: true
    include-in-header: 
      - metadata.tex
      - header.tex
    include-before-body: title.tex
execute:
  warning: false
  message: false
editor: visual
---

```{r}
#| echo: false

# Load data: 
require(ElemStatLearn)
data('SAheart')
# View(SAheart)

# Random training data set:
set.seed(1)
training = SAheart[sample(1:462,size=300),]
```

```{r}
#| echo: false

# load required libraries:
library(betareg)
library(ggplot2)
library(dplyr)
library(purrr)
```

## Journal Entry 1

17/08/2026

During this lecture I understood most of what was discussed as we went through notation and basic time series components. I do want to practice plotting some of the built in R datasets to see how they look. I want to see how a qualitative, continuous and discrete time series looks.

For a Qualitative:

```{r}
library(ctsfeatures)
data("SleepStages")

library(dplyr)
library(lubridate)

# Aggregate the data from seconds to minutes
SleepStages_min <- SleepStages |>
  mutate(
    Time2 = hms(as.character(Time)),
    Minute = floor(period_to_seconds(Time2) / 60)
  ) |>
  group_by(Series, Minute) |>
  count(Value) |>
  slice_max(n, n = 1, with_ties = FALSE) |>
  select(Series, Minute, Value)

# Choose one series first
sleep1 <- SleepStages |>
  filter(Series == 1) |>
  mutate(
    Time_sec = period_to_seconds(hms(as.character(Time))),
    Time_min = Time_sec / 60
  )

ggplot(sleep1, aes(x = Time_min, y = Value, group = 1)) +
  geom_step(col = "hotpink", lwd = 0.7) +
  labs(
    title = "Sleep Stages over Time",
    x = "Time (minutes)",
    y = "Sleep Stage"
  ) +
  theme_minimal()
```

For Continuous:

```{r}
data(Nile)

plot.ts(Nile,
        main = "Plot of the Nile dataset.",
        xlab = "Year",
        ylab = "Flow",
        col = "hotpink",
        type = "l",
        lwd = 2,
        lty = 1
        )
```

For discrete:

```{r}
data(lynx)

barplot(
  as.numeric(lynx),
  main = "Annual Canadian Lynx Trappings",
  xlab = "Year",
  ylab = "Number of Lynx",
  col = "hotpink"
)
```

Now I will follow the same structure as in the "Time Series Analysis in R.pdf" and practice some of the functions taught in class.

Aggregation example:

```{r}
data("AirPassengers")

print(frequency(AirPassengers))

months_to_years <- aggregate(AirPassengers, nfrequency = 1, FUN = mean)
class(months_to_years)

plot.ts(months_to_years,
        main = "Plot of the aggrgated AirPassengers data.",
        xlab = "Year",
        ylab = "Yearly Mean.",
        col = "hotpink",
        type = "l",
        lwd = 2,
        lty = 1
)
```

Since the AirPassenger data is monthy (frequency = 12) I aggregated it to annual data to see what it would look like since we are lower the granularity.

Window:

Now I want to plot a subset of the AirPassenger data to practice writing out the window function.

```{r}
window1 <- window(AirPassengers, start = c(1955, 1), end = c(1958, 12))

plot.ts(window1,
        main = "Plot of the a window of the AirPassengers data.",
        xlab = "Month",
        ylab = "Total airline passengers.",
        col = "hotpink",
        type = "l",
        lwd = 2,
        lty = 1
)
```

Trying lag() and diff() on a simulated dataset:

```{r}
ts_sim <- ts(c(11, 16, 21, 33, 40, 54, 59, 60, 66, 70, 73, 75),
             start = c(2025, 1),
             frequency = 12)

print(ts_sim)

# Shifts the time index backwards by 1 period
lagged_ts <- stats::lag(ts_sim, k = 1)
lagged_ts

# First order differencing removes the linear trend
diff_ts <- diff(ts_sim, differences = 1)
diff_ts
```

Checking out ts.union() and ts.intersect():

```{r}
ts_a <- ts(1:10, start = c(2023, 1), frequency = 12)
ts_b <- ts(11:18, start = c(2023, 4), frequency = 12)

# Union
combined_un <- ts.union(ts_a, ts_b)
combined_un

# Intersect
itersect_ts <- ts.intersect(ts_a, ts_b)
itersect_ts
```

What is the structure of the ts object:

```{r}
raw_data <- data.frame(
  Month = 1:12,
  Value = c(10, 12, 15, 14, 18, 22, 25, 24, 20, 15, 12, 11,
            12, 14, 17, 16, 20, 24, 28, 27, 22, 17, 14, 13)
)

ts_sim2 <- ts(raw_data$Value, start = c(2024, 1), frequency = 12)

decom <- decompose(ts_sim2)
plot(decom)

plot(ts_sim2, type = "o", pch = 20, main = "Demonstration of ts Object.", ylab = "Values", xlab = "Time", col = "hotpink")
```

## Journal Entry 2

19/08/2026

Today we went over time series cross validation, below I will attempt to construct an expanding-window cross validation from first principles:

```{r}
library(forecast)

y <- AirPassengers

n <- length(y)

initial <- 24

errors <- rep(NA, n)
predictions <- rep(NA, n)

for (T in initial:(n - 1)) {
  
  # 1) Create training data
  train <- y[1:T]
  
  # 2) Seasonal naive forecast
  # same month last year
  prediction <- train[T - 12 + 1]
  
  # 3) Get actual future observation
  actual <- y[T + 1]
  
  # 4) Calculate forecast error
  error <- actual - prediction
  
  # 5) Store results
  predictions[T + 1] <- prediction
  errors[T + 1] <- error
}

RMSE <- sqrt(mean(errors^2, na.rm = TRUE))

RMSE

MAE <- mean(abs(errors), na.rm = TRUE)

MAE
```

Which holds since RMSE \> MAE.

Exercises slide 72:

1\) Why is ordinary random k-fold cross-validation inappropriate for forecasting?

Ordinary random k-fold cross-validation divides the data into k folds and randomly leaves one fold out for validating the data. The remaining k -1 folds are used as the training set and the model is trained with those. The issue is that Time Series data is sequential and dependent and randomly splitting the dataz into k-folds would destroy the temporal ordering of the series So the test set can only be values that are ahead of the training values.

2\) Define temporal leakage.

Temporal leakage is when data from the future is used to train a model that has to predict that future.

3\) Contrast Walk-Forward, Expanding Window and Sliding Window Validation.

4\) When would Sliding window outperform Expanding window?

I think that sliding window will outperform the expanding window when beyond a certain past time threshold the data will not be informative for forecasting the future. For example milk sales data from 30 years ago might not apply to milk sales today with people changing to plant based milks etc so the market trends might not be applicable now.

5\) A firm refits its model daily using only the most recent 250 observations. Which strategy is this?

This is the sliding window strategy, because the model is being refitted to a fixed sized training window containing only the most recent 250 observations.

6\) Why must temporal ordering be preserved during validation?

Because otherwise the results will not be accurate as Time Series data must be temporal.
