---
title: "Section 1 Rough"
author: "Roxy Campher"
format: pdf
editor: visual
---

## Rough working of section 1

A QMD for playing around with base R time series datasets. I want to explore how to fit the models on a basic dataset before diving into more complex data.

### Journal notes

Make a journal entry about exploring basic Time Series datasets.

## Base R Time Series Datasets

### AirPassengers

```{r}
data(AirPassengers)

AirPassengers
class(AirPassengers) # it is a ts object

plot(AirPassengers)

frequency(AirPassengers) # number of observations per unit
start(AirPassengers) # returns the start time of the series
end(AirPassengers) # returns the end time of the series

acf(AirPassengers)
pacf(AirPassengers)

win <- window(AirPassengers, start = c(1950, 1), end = c(1955, 12))
print(win)

diff_delta <- diff(AirPassengers)
plot(diff_delta)
```

### Nile - ARIMA

Annual - no seasonal component to worry about.

```{r}
data(Nile)
plot(Nile)
```

### Nottem

Has obvious seasonality

```{r}
data("nottem")
plot(nottem)
```

### EUStockMarkets

This is particularly useful for understanding the difference between a nonstationary price/index series and approximately stationary returns.

```{r}
data("EuStockMarkets")
plot(EuStockMarkets)

dax <- EuStockMarkets[, "DAX"]
plot(dax)

returns <- diff(log(dax))
plot(returns)
```

# Portfolio models to compare with market volatility data - JSE?

Ask Tim for data or where to download data.

ARCH and GARCH vs Auto regressive models. Particularly - ARIMA vs GARCH could be useful.

ARIMA - behaviour

GARCH - variance/volatility behaviour
