---
title: "Lecture 1 and 2 Workthrough"
author: "Roxy Campher"
format: pdf
editor: visual
---

```{r}
#| echo: false

# load required libraries:
library(betareg)
library(ggplot2)
library(dplyr)
library(purrr)
```

For a Qualitative Time Series:

```{r}
library(ctsfeatures)
data("SleepStages")

library(dplyr)
library(lubridate)

# Aggregate the data from seconds to minutes
SleepStages_min <- SleepStages |>
  mutate(
    Time2 = hms(as.character(Time)),
    Minute = floor(period_to_seconds(Time2) / 60)
  ) |>
  group_by(Series, Minute) |>
  count(Value) |>
  slice_max(n, n = 1, with_ties = FALSE) |>
  select(Series, Minute, Value)

# Choose one series first
sleep1 <- SleepStages |>
  filter(Series == 1) |>
  mutate(
    Time_sec = period_to_seconds(hms(as.character(Time))),
    Time_min = Time_sec / 60
  )

ggplot(sleep1, aes(x = Time_min, y = Value, group = 1)) +
  geom_step(col = "hotpink", lwd = 0.7) +
  labs(
    title = "Sleep Stages over Time",
    x = "Time (minutes)",
    y = "Sleep Stage"
  ) +
  theme_minimal()
```

For Continuous Time Series:

```{r}
data(Nile)

plot.ts(Nile,
        main = "Plot of the Nile dataset.",
        xlab = "Year",
        ylab = "Flow",
        col = "hotpink",
        type = "l",
        lwd = 2,
        lty = 1
        )
```

For Discrete Time Series:

```{r}
data(lynx)

barplot(
  as.numeric(lynx),
  main = "Annual Canadian Lynx Trappings",
  xlab = "Year",
  ylab = "Number of Lynx",
  col = "hotpink"
)
```

Now I will follow the same structure as in the "Time Series Analysis in R.pdf" and practice some of the functions taught in class.

Aggregation example:

```{r}
data("AirPassengers")

print(frequency(AirPassengers))

months_to_years <- aggregate(AirPassengers, nfrequency = 1, FUN = mean)
class(months_to_years)

plot.ts(months_to_years,
        main = "Plot of the aggrgated AirPassengers data.",
        xlab = "Year",
        ylab = "Yearly Mean.",
        col = "hotpink",
        type = "l",
        lwd = 2,
        lty = 1
)
```

Since the AirPassenger data is monthly (frequency = 12) I aggregated it to annual data to see what it would look like since we are lower the granularity.

Window:

Now I want to plot a subset of the AirPassenger data to practice writing out the window function.

```{r}
window1 <- window(AirPassengers, start = c(1955, 1), end = c(1958, 12))

plot.ts(window1,
        main = "Plot of the a window of the AirPassengers data.",
        xlab = "Month",
        ylab = "Total airline passengers.",
        col = "hotpink",
        type = "l",
        lwd = 2,
        lty = 1
)
```

Trying lag() and diff() on a simulated dataset:

```{r}
ts_sim <- ts(c(11, 16, 21, 33, 40, 54, 59, 60, 66, 70, 73, 75),
             start = c(2025, 1),
             frequency = 12)

print(ts_sim)

# Shifts the time index backwards by 1 period
lagged_ts <- stats::lag(ts_sim, k = 1)
lagged_ts

# First order differencing removes the linear trend
diff_ts <- diff(ts_sim, differences = 1)
diff_ts
```

Checking out ts.union() and ts.intersect():

```{r}
ts_a <- ts(1:10, start = c(2023, 1), frequency = 12)
ts_b <- ts(11:18, start = c(2023, 4), frequency = 12)

# Union
combined_un <- ts.union(ts_a, ts_b)
combined_un

# Intersect
itersect_ts <- ts.intersect(ts_a, ts_b)
itersect_ts
```

What is the structure of the ts object:

```{r}
raw_data <- data.frame(
  Month = 1:12,
  Value = c(10, 12, 15, 14, 18, 22, 25, 24, 20, 15, 12, 11,
            12, 14, 17, 16, 20, 24, 28, 27, 22, 17, 14, 13)
)

ts_sim2 <- ts(raw_data$Value, start = c(2024, 1), frequency = 12)

decom <- decompose(ts_sim2)
plot(decom)

plot(ts_sim2, type = "o", pch = 20, main = "Demonstration of ts Object.", ylab = "Values", xlab = "Time", col = "hotpink")
```
